Model Adaptivity for Goal-Oriented Inference using Adjoints

### 2017-06-15 paper/submitted/round2_png/mathematical_formulation_v4_cmame.tex
Expanded proposition 1. amd turned Eq. 14 back into a proposition

Let $\Psi_{LF}$ denote the stationary point of the low-fidelity augmented Lagrangian $\mathcal{M}_{LF}$ induced by~\Cref{eq:invOptLF}, respectively. Then, the error in the Quantity of Interest $I$ is given by
%
\begin{multline}
\label{eq:semifinErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{,}
\end{multline}
%
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$.
\end{proposition}
%
\begin{proof}
%
Observe that
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

Extending the property in \Cref{eq:MeqI} to the augmented Lagrangians for the high- and low-fidelity models, we have
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Applying the output error representation described in Proposition 3 from~\cite{BecVex05} for the difference $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$,
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}(e^3)\textrm{.}
\end{equation}
Combining \Cref{eq:repIwithM} and \Cref{eq:beckvex} we obtain
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}(e^3)\textrm{,}
\end{multline}
which completes the proof.
\end{proof}
%

The above proposition splits the error in the QoI into three components, a third order remainder term $\mathcal{R}(e^3)$, a computable `bias' term $\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})$, and a linear term $\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})$. Although the linear term is not computable, it can be estimated using an appropriate dual problem. We describe one method of estimating this term below.

Note that, as a stationary point, $\Psi_{HF}$ satisfies~\Cref{eq:supadjsys}. Now, let $\mathcal{Q}$ be an output defined by
%
\begin{equation}
\mathcal{Q}(\Phi)=\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Phi).
\label{eq:supadjout}
\end{equation}
%
Let the variational form
%
\begin{equation}
\mathscr{R}(\Psi_{HF})(\Phi)=0,\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:supadjsys}
\end{equation}
%
represent the equation $\mathcal{M}'_{HF,\Psi}(\Psi_{HF})(\Phi)=0$ in residual form, where $\Phi$ is a test function. Define the corresponding adjoint problem
%
\begin{equation}
\mathscr{R}_{\Psi}'(\Psi_{HF},\Phi)(\Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:superAdjEq}
\end{equation}
for the supplementary adjoint $\Lambda$, where $\mathscr{R}$ is defined in \Cref{eq:supadjsys}. The error in the output $\mathcal{Q}$ defined in \Cref{eq:supadjout} can then be expressed as a dual-weighted residual,
\begin{equation}
\label{eq:adjOutErr}
\mathcal M'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})=-\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda).
\end{equation}
%
Combining \Cref{eq:preadj} and \Cref{eq:adjOutErr}, we have,
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF})+\mathcal{R}(e^3).
\end{multline}
%
Other approaches can become feasible in certain scenarios; for example, if the high- and low-fidelity models only differ in the computational grid used, the difference $\Psi_{HF}-\Psi_{LF}$ can be approximated using interpolation or patch recovery methods~\cite{BecVex05}.

>>>>

Let $\Psi_{LF}$ denote the stationary point of the low-fidelity augmented Lagrangian $\mathcal{M}_{LF}$ induced by~\Cref{eq:invOptLF}, respectively. Then, the error in the Quantity of Interest $I$ is given by
%
\begin{multline}
\label{eq:semifinErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}\textrm{,}
\end{multline}
%
where $\mathcal{R}$ is a remainder term that is third-order in the error $e=\Psi_{HF}-\Psi_{LF}$.
\end{proposition}
%
\begin{proof}
%
Observe that
%
\begin{equation}
\label{eq:MeqI}
\mathcal{M}(\Psi)=I(q,u),
\end{equation}
%
since taking variations of $\mathcal{M}$ with respect to the auxiliary variables gives that $\xi_\Psi$ is a stationary point of $\mathcal{L}$.

Extending the property in \Cref{eq:MeqI} to the augmented Lagrangians for the high- and low-fidelity models, we have
%
\begin{multline}
\label{eq:repIwithM}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})\textrm{.}
\end{multline}
%
Following~\cite{BecVex05}, we rewrite the term $\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF})$ using the Fundamental Theorem of Calculus,
%
\begin{equation}
\label{eq:Mantiderivative}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \int\limits_{0}^{1} \mathcal{M}'_{HF}\left(\Psi_{HF} + se\right)\left(e\right) \, ds
\end{equation}
%
Now using the trapezoidal rule for the integral in~\Cref{eq:Mantiderivative}, we further have,
%
\begin{equation}
\label{eq:Mprimetrapezoid}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{HF})(e) + \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(e) + \mathcal{R}\textrm{.}
\end{equation}
%
where the remainder $R$ is given as,
%
\begin{equation}
\label{eq:trapezoidremainder}
R = \frac{1}{2} \int\limits_{0}^{1} \mathcal{M}'''_{HF}\left(\Psi_{LF} + se\right)(e,e,e) \, s(s-1) \, ds
\end{equation}
%
which is finite due to the third order differentiability assumptions in the Proposition. We further note that $\mathcal{M}'_{HF}(\Psi_{HF})(e)$ vanishes since $\Psi_{HF}$ is a stationary point of $\mathcal{M}_{HF}$, and~\Cref{eq:Mprimetrapezoid} can be simplified to,
%
\begin{equation}
\label{eq:beckvex}
\mathcal{M}_{HF}(\Psi_{HF})-\mathcal{M}_{HF}(\Psi_{LF}) = \frac{1}{2}\mathcal{M}'_{HF}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{R}\textrm{.}
\end{equation}
%
Combining \Cref{eq:repIwithM} and \Cref{eq:beckvex} we obtain,
%
\begin{multline}
\label{eq:preadj}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})+\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})+\mathcal{R}\textrm{,}
\end{multline}
which completes the proof.
\end{proof}
%

The above proposition splits the error in the QoI into three components, a third order remainder term $\mathcal{R}$, a computable `bias' term $\mathcal{M}_{HF}(\Psi_{LF})-\mathcal{M}_{LF}(\Psi_{LF})$, and a linear term $\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Psi_{HF}-\Psi_{LF})$. Although the linear term is not computable, it can be estimated. We describe one method of estimating this term below, using an appropriate dual problem.

%
\begin{proposition}
\label{thm:error_estimate_dual}
Let the hypotheses of Proposition~\ref{thm:error_estimate} hold. Let $Lambda$ be the solution of the dual problem,
%
\begin{equation}
\mathcal{M}_{HF}'(\Psi_{LF};\Phi, \Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\end{equation}
%
where $\mathcal{Q}:(U \times U \times Q)^2 \to \R$ is a linear functional defined by,
%
\begin{equation}
\mathcal{Q}(\Phi)=\mathcal{M}'_{HF}(\Psi_{LF})(\Phi).
\label{eq:supadjout}
\end{equation}
%
Then we have,
%
\begin{multline}
\label{eq:finErrExp}
I(q_{HF},u_{HF})-I(q_{LF},u_{LF})=\\-\frac{1}{2}\mathcal{M}'_{HF,\Psi}(\Psi_{LF})(\Lambda)+\mathcal M_{HF}(\Psi_{LF})-\mathcal M_{LF}(\Psi_{LF}) + R_{\mathscr{R}} + R
\end{multline}
%
where $R_{\mathscr{R}}$ is a second order term in the error $e=\Psi_{HF}-\Psi_{LF}$, and $R$ is as defined in Proposition~\ref{thm:error_estimate}.
%
\end{proposition}
%
\begin{proof}
%
The proof simply follows from the classic dual based a posteriori error estimation strategy. Note that, as a stationary point, $\Psi_{HF}$ satisfies~\Cref{eq:supadjsys}. Let the variational form
%
\begin{equation}
\mathscr{R}(\Psi_{HF},\Phi)=0,\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:supadjsys}
\end{equation}
%
represent the equation $\mathcal{M}'_{HF}(\Psi_{HF})(\Phi)=0$ in residual form, where $\Phi$ is a test function. Define the corresponding adjoint problem
%
\begin{equation}
\mathscr{R}'(\Psi_{LF};\Phi,\Lambda)=\mathcal{Q}(\Phi),\quad\forall\Phi\in(Q_{HF}\times U_{HF}\times U_{HF})^2
\label{eq:superAdjEq}
\end{equation}
%
for the supplementary adjoint $\Lambda$. The error in the linear output $\mathcal{Q}$ defined in \Cref{eq:supadjout} can then be expressed as a dual-weighted residual,
%
\begin{equation}
\label{eq:adjOutErr}
\mathcal{Q}(\Psi_{HF}-\Psi_{LF})=-\mathscr{R}(\Psi_{LF},\Lambda) + R_{\mathscr{R}}
\end{equation}
%
where $R_{\mathscr{R}}$ is a second order remainder term due to the linearization of $\mathscr{R}$ about $\Psi_{LF}$. This implies the required result.
%
\end{proof}
%
Other approaches can become feasible in certain scenarios; for example, if the high- and low-fidelity models only differ in the computational grid used, the difference $\Psi_{HF}-\Psi_{LF}$ can be approximated using interpolation or patch recovery methods~\cite{BecVex05}.
